---
title: "Beyond the Finish Line: A Glance at Formula 1 and Ferrari"
author: "Eric L. Rehana H. Fatima T."
date: "05-06-2025"
output:
  html_document:
    code_folding: show
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE
    
    
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


# Introduction
#### Add intro to project here


# What We Were Looking For
#### Add questions here


# Packages Used
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
```

# Our Dataset
##### [Our Data](https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020)
##### Include info from the our dataset section on our ppt here


# Prepping our Data
##### Include info on how we prepped our data


# A Glance at F1
##### Include general graphs and info here

# Let's Take a Look at Ferrari!
##### Include Ferrari graphs one chunk/graph at a time and discuss findings 


# Comparing Ferrari and McLaren 
##### Info here

# Our Limitations
##### Infor here

# Conclusion
##### Info here

