---
title: "Beyond the Finish Line: A Glance at Formula 1 and Ferrari"
author: "Eric L. Rehana H. Fatima T."
date: "05-06-2025"
output:
  html_document:
    code_folding: show
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

#### Add intro to project here

# What We Were Looking For

#### Add questions here

# Packages Used

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
```

# Our Dataset

##### [Our Data](https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020)

##### Include info from the our dataset section on our ppt here

### The CSV files

```{r}
#reading files
circuits<- read.csv("f1data/circuits.csv")
constructor_results<- read.csv("f1data/constructor_results.csv")
constructor_standings<- read.csv("f1data/constructor_standings.csv")
constructors<- read.csv("f1data/constructors.csv")
driver_standings<- read.csv("f1data/driver_standings.csv")
drivers<- read.csv("f1data/drivers.csv")
lap_times<- read.csv("f1data/lap_times.csv")
pit_stops<- read.csv("f1data/pit_stops.csv")
qualifying<- read.csv("f1data/qualifying.csv")
races<- read.csv("f1data/races.csv")
results<- read.csv("f1data/results.csv")
seasons<- read.csv("f1data/seasons.csv")
sprint_results<- read.csv("f1data/sprint_results.csv")
status<- read.csv("f1data/status.csv")

```

# Prepping our Data

##### Include info on how we prepped our data

### Functions and mutations

```{r}
#Functions

#convert "min:sec.ms" string to total seconds (as a number)
time_to_seconds <- function(time_str) {
  parts <- str_split_fixed(time_str, ":", 2)
  minutes <- as.numeric(parts[, 1])
  seconds_ms <- as.numeric(parts[, 2])
  return(minutes * 60 + seconds_ms)
}

#convert total seconds back to "min:sec.ms" format
seconds_to_time <- function(total_seconds) {
  minutes <- floor(total_seconds / 60)
  remaining_seconds <- total_seconds - (minutes * 60)
  return(sprintf("%d:%.3f", minutes, remaining_seconds))
}

#average lap times and mutate a new column
qualifying_average <- qualifying %>%
  mutate(
    q1_seconds = sapply(q1, time_to_seconds),
    q2_seconds = sapply(q2, time_to_seconds),
    q3_seconds = sapply(q3, time_to_seconds),
    average_seconds = case_when(
      !is.na(q1_seconds) & !is.na(q2_seconds) & !is.na(q3_seconds) ~ (q1_seconds + q2_seconds + q3_seconds) / 3,
      !is.na(q1_seconds) & !is.na(q2_seconds) & is.na(q3_seconds) ~ (q1_seconds + q2_seconds) / 2,
      TRUE ~ q1_seconds 
    ),
    average_time = sapply(average_seconds, seconds_to_time)
  ) %>%
  select(-q1_seconds, -q2_seconds, -q3_seconds)
```

### Joins

```{r}
#Joins to add years to data frames for graphing

#qualifying data
qualifying_by_year<-left_join(races, qualifying_average) 
qualifying_by_year = left_join(qualifying_by_year, constructors, by = join_by(constructorId))
#delete
qualifying_by_year$name.y = as.character(qualifying_by_year$name.y)

#results data by year
results_LapTimeSeconds <- results %>%
  mutate(fastestLapTime_seconds = sapply(fastestLapTime , time_to_seconds))

results_by_year<- left_join(races, results_LapTimeSeconds, by = join_by(raceId))
results_by_year = left_join(results_by_year, constructors, by = join_by(constructorId))

#constructor standings by year
constructor_standings_by_year<-left_join(races, constructor_standings)
```

# A Glance at F1

### Filtering and color coordinating the current teams on the grid

```{r}
#filters for current teams on the grid
relevant_results = left_join(x = constructor_standings_by_year, y = constructors, 
                  by = join_by(constructorId)) %>% filter(name.y == 
                  'Alpine F1 Team' | name.y == 'Aston Martin' | name.y == 'Ferrari'| 
                          name.y =='Haas F1 Team' | name.y == 'McLaren' | 
                          name.y == 'Mercedes' | name.y == 'AlphaTauri' |
                          name.y == 'Red Bull' | name.y == 'Sauber' |
                          name.y == 'Williams')

#setting color palette
colorpalette = c('#041F30', '#FD4BC7', '#00352F', '#A6051A', '#AEAEAE', '#FF8000', 
'#111111', '#223971', '#00E701', '#00A3E0')
names(colorpalette) = c('Alpine F1 Team', 'Aston Martin', 'Ferrari', 'Haas F1 Team', 
                 'McLaren', 'Mercedes', 'AlphaTauri', 'Red Bull', 'Sauber', 'Williams')
```

##### Include general graphs and info here

```{r}
#Graphs for Part A: Overview of all current F1 teams

#average qual time by year
qualifying_by_year %>% filter(name.y == 'Alpine F1 Team' | 
                      name.y == 'Aston Martin' | name.y == 'Ferrari' | 
                      name.y =='Haas F1 Team' | name.y == 'McLaren' | 
                      name.y == 'Mercedes' | name.y == 'AlphaTauri' |
                      name.y == 'Red Bull' | name.y == 'Sauber' |
                      name.y == 'Williams') %>% ggplot(aes(x = year,
                                                y = average_seconds, color = 
                                                name.y)) + geom_smooth(se = 
                                                          FALSE)+labs(
                          title = "Average Qualifying Times by Season",
          subtitle = "All current teams",
          x = "Season", y = "Average Qualifying Time (min:ss)", 
          color = 'Teams')  +   theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    legend.background = element_rect(fill = "black")
  )

#wins by year for current teams
overview_results = left_join(x =races, y = relevant_results, by = join_by(raceId))
overview_results = left_join(overview_results, constructors, by = join_by(constructorId))
constructor_wins = overview_results %>%  group_by(name.y, year.x) %>% 
  summarise(total_wins = max(wins, na.rm = TRUE)) %>% drop_na(name.y) 


ggplot(constructor_wins, aes(x = year.x, y = total_wins, fill = name.y)) +
  geom_col(color = '#616161', width = 2, alpha = 1/2) + labs(title = 
                                          "Wins by Season",
                                          subtitle = "All current teams",
                                         x = "Season", y = "Wins", 
                                         fill = 'Teams') + 
   theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    legend.background = element_rect(fill = "black")
  )




#fastest lap time by year
results_by_year %>% filter(name.y == 'Alpine F1 Team' | 
                      name.y == 'Aston Martin' | name.y == 'Ferrari' | 
                      name.y =='Haas F1 Team' | name.y == 'McLaren' | 
                      name.y == 'Mercedes' | name.y == 'AlphaTauri' |
                      name.y == 'Red Bull' | name.y == 'Sauber' |
                      name.y == 'Williams') %>% drop_na(fastestLapTime_seconds) %>% ggplot(aes(x = year, y = 
                                              fastestLapTime_seconds, color = name.y)) + 
                                              geom_smooth(alpha = 1/5, se = FALSE) + labs(title = 
                                          "Fastest Lap Times by Season",
                                          subtitle = "All current teams",
                                         x = "Season", y = "Lap Time (min:ss)", 
                                         color = 'Teams') + 
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    legend.background = element_rect(fill = "black")
  )

```

# Let's Take a Look at Ferrari!

### Filtering for Ferrari data and averaging data per season

```{r}
# graphs for Part B: A closer look at Ferrari

#filtering for Ferrari
ferrari_qualifying<- qualifying_by_year %>% filter(constructorId == 6)
ferrari_results<- results_by_year %>% filter(constructorId == 6)
ferrari_standings<- constructor_standings_by_year %>% filter(constructorId ==6)

#summarizing for average data per season ferrari
ferrari_avg_qualifyingTime<- ferrari_qualifying %>% group_by(year) %>% summarise(avg_time_per_season = mean(average_seconds, na.rm = TRUE))

ferrari_avg_LapTime<- ferrari_results %>% group_by(year) %>% summarise(avg_fastestLapTime_per_season = mean(fastestLapTime_seconds, na.rm = TRUE)) %>% filter(!is.na(avg_fastestLapTime_per_season))

ferrari_avg_fastestLapSpeed<- ferrari_results %>% group_by(year) %>% summarise(avg_fastestLapSpeed_per_season = mean(as.numeric(fastestLapSpeed), na.rm = TRUE)) %>% filter(!is.na(avg_fastestLapSpeed_per_season))

ferrari_avg_finishPosition<- ferrari_results %>% group_by(year) %>% summarise(avg_finishPosition_per_season = mean(positionOrder, na.rm = TRUE)) %>% filter(!is.na(avg_finishPosition_per_season))

ferrari_avg_rank<- ferrari_results %>% group_by(year) %>% summarise(avg_rank_per_season = mean(as.numeric(rank), na.rm = TRUE)) %>% filter(!is.na(avg_rank_per_season))

ferrari_avg_runTime<- ferrari_results %>% group_by(year) %>% summarise(avg_runTime_per_season = mean(as.numeric(milliseconds), na.rm = TRUE)) %>% filter(!is.na(avg_runTime_per_season))

ferrari_total_wins<- ferrari_standings %>% group_by(year) %>% summarise(total_wins = max(wins, na.rm = TRUE))

ferrari_avg_qualifyingPos<- ferrari_qualifying %>% group_by(year) %>% summarise(avg_pos_per_season = mean(position, na.rm = TRUE))
```

##### Include Ferrari graphs one chunk/graph at a time and discuss findings

```{r}
#ferrari qualifying times by year
ferrari_avg_qualifyingTime %>% 
  ggplot(aes(x = year, y = avg_time_per_season)) + 
  geom_point(color = '#FFF200') + 
  geom_smooth(color = '#EF1A2D') + 
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') + 
  labs(
    title = "Ferrari Average Qualifying Times by Season",
    subtitle = "Measured in min:ss | Lower is better | From 1994 to 2024",
    x = "Season",
    y = "Average Qualifying Time"
  )+ 
  scale_x_continuous(
    limits = c(min(ferrari_avg_qualifyingTime$year), max(ferrari_avg_qualifyingTime$year)),
    expand = c(0, 0)
  )  + 
  scale_y_continuous(
    labels = function(x) paste0(sprintf("%02d", floor(x / 60)), ":", sprintf("%02d", round(x %% 60))) 
  ) + 
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


```{r}
#Ferrari fastest lap times by year
ferrari_avg_LapTime %>% 
  ggplot(aes(x = year, y = avg_fastestLapTime_per_season)) + 
  geom_point(color = '#FFF200') + 
  geom_smooth(color = '#EF1A2D') + 
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') + 
  labs(
    title = "Ferrari Fastest Lap Times by Season",
    subtitle = "Measured in min:ss | Lower is better | From 2004 - 2024",
    x = "Season",
    y = "Average Fastest Lap Time"
  ) + 
  scale_x_continuous(
    limits = c(2005, 2025),
    expand = c(0, 0.5)
  ) + 
  scale_y_continuous(
    labels = function(x) paste0(sprintf("%02d", floor(x / 60)), ":", sprintf("%02d", round(x %% 60)))
  ) + 
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


```{r}
#Ferrari fastest lap speed over time
ferrari_avg_fastestLapSpeed %>%
  ggplot(aes(x = year, y = avg_fastestLapSpeed_per_season)) +
  geom_point(color = '#FFF200') +
  geom_smooth(color = '#EF1A2D') +
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') +
  labs(
    title = "Ferrari Fastest Lap Speed by Season",
    subtitle = "Measured in km/h and mph | From 2004 to 2024",
    x = "Season"
  ) +
  scale_y_continuous(
    name = "Average Fastest Lap Speed (km/h)",
    sec.axis = sec_axis(~ . * 0.621371, name = "Average Fastest Lap Speed (mph)")
  ) +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    axis.text.y.right = element_text(color = "white"),
    axis.title.y.right = element_text(color = "white"), 
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


```{r}
#Ferrari finish position over time
ferrari_avg_finishPosition %>%
  ggplot(aes(x = year, y = avg_finishPosition_per_season)) +
  geom_point(color = '#FFF200') +
  geom_smooth(color = '#EF1A2D') +
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') +
  labs(
    title = "Ferrari Average Finish Position by Season",
    subtitle = "Lower is better | From 1950 to 2024",
    x = "Season",
    y = "Average Finish Position"
  ) +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


```{r}
#Ferrari rank over time
ferrari_avg_rank %>%
  ggplot(aes(x = year, y = avg_rank_per_season)) +
  geom_point(color = '#FFF200') +
  geom_smooth(color = '#EF1A2D') +
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') +
  labs(
    title = "Ferrari Average Rank by Season",
    subtitle = "Lower is better | From 2004 to 2024",
    x = "Season",
    y = "Average Finish Rank"
  ) + 
  scale_x_continuous(
    limits = c(min(ferrari_avg_rank$year), max(ferrari_avg_rank$year) + 1),
    expand = c(0, .5) 
  ) + 
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


```{r}
#Ferrari race run time over time
ferrari_avg_runTime %>%
  ggplot(aes(x = year, y = as.numeric(avg_runTime_per_season)/3600000)) +
  geom_point(color = '#FFF200') +
  geom_smooth(color = '#EF1A2D') +
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') +
  labs(
    title = "Ferrari Average Runtime by Season",
    subtitle = "Converted from milliseconds to hours | From 1950 to 2024",
    x = "Season",
    y = "Average Runtime (Hours)"
  ) +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


```{r}
#Ferrari wins over time
ferrari_total_wins %>% ggplot(aes(x = year, y = total_wins)) + geom_point(color = '#FFF200') + geom_line(color = '#EF1A2D') + geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') + theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold"),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  ) + scale_y_continuous(breaks = seq(0, max(ferrari_total_wins$total_wins), by = 1)) + scale_x_continuous(breaks = seq(1955, 2025, by = 5)) + labs(
    title = "Ferrari Wins by Season",
    subtitle = "From 1958 to 2024",
    x = "Season",
    y = "Total Wins"
  )
```


```{r}
#Graph for Ferrari qualifying position over time 
ferrari_avg_qualifyingPos %>% 
  ggplot(aes(x = year, y = avg_pos_per_season)) + 
  geom_point(color = '#FFF200') + 
  geom_smooth(color = '#EF1A2D') + 
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') + 
  labs(
    title = "Ferrari Average Qualifying Position by Season",
    subtitle = "Lower is better | From 1994 to 2024",
    x = "Season",
    y = "Average Qualifying Position"
  )+ 
  scale_x_continuous(
    limits = c(min(ferrari_avg_qualifyingPos$year), max(ferrari_avg_qualifyingPos$year)),
    expand = c(0, 0)
  ) + theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


```{r}
#Graph to compare Ferrari qualifying position to finish position scaled to match lesser time frame
ferrari_avg_finishPosition %>%
  ggplot(aes(x = year, y = avg_finishPosition_per_season)) +
  geom_point(color = '#FFF200') +
  geom_smooth(color = '#EF1A2D') +
  geom_smooth(linetype = "longdash", method = lm, se = FALSE, color = '#00A551') +
  scale_x_continuous(limits = c(1994, 2024), breaks = seq(1994, 2024, by = 5)) +
  scale_y_continuous(limits = c(2.5, 10)) +
  labs(
    title = "Ferrari Average Finish Position by Season",
    subtitle = "Lower is better | From 1994 to 2024",
    x = "Season",
    y = "Average Finish Position"
  ) +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid.major = element_line(color = "gray30"),
    panel.grid.minor = element_line(color = "gray20"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", face = "bold", size = 16),
    plot.subtitle = element_text(color = "white", size = 12, margin = margin(b = 10)),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white")
  )
```


# Comparing Ferrari and McLaren

##### Info here

# Our Limitations

##### Info here

# Conclusion

##### Info here
